\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-03-03 analysis/paper/paper.tex

The sampling scheme selected for \code{sampleCaseBase} was described in
Hanley and Miettinen \citeyearpar{hanley2009fitting}: we first sample
along the ``person'' axis, proportional to each individual's total
follow-up time, and then we sample a moment uniformly over their
follow-up time. This sampling scheme is equivalent to the following
picture: imagine representing the total follow-up time of all
individuals in the study along a single dimension, where the follow-up
time of the next individual would start exactly when the follow-up time
of the previous individual ends. Then the base series could be sampled
uniformly from this one-dimensional representation of the overall
follow-up time. In any case, the output is a dataset of the same class
as the input, where each row corresponds to a person-moment. The
covariate profile for each such person-moment is retained, and an offset
term is added to the dataset. This output could then be used to fit a
smooth hazard function, or for visualization of the base series.

The fitting function \code{fitSmoothHazard} starts by looking at the
class of the dataset: if it was generated from \code{sampleCaseBase}, it
automatically inherited the class \code{cbData}. If the dataset supplied
to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
fitting function starts by calling \code{sampleCaseBase} to generate the
base series. In other words, the occasional user can bypass
\code{sampleCaseBase} altogether and only worry about the fitting
function \code{fitSmoothHazard}.

The fitting function retains the familiar formula interface of
\code{glm}. The left-hand side of the formula should be the name of the
column corresponding to the event type. The right-hand side can be any
combination of the covariates, along with an explicit functional form
for the time variable. Note that non-proportional hazard models can be
achieved at this stage by adding an interaction term involving time. The
offset term does not need to be specified by the user, as it is
automatically added to the formula.

@@

\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
From this, we can derive the event-specific subdensities:

\[ f_j(t) = h_j(t)S(t).\]

@@

\[ CI_j(t) = \int_0^t f_j(u)du.\]

We created \code{absoluteRisk} as an \code{S3} generic, with methods for
the different types of outputs of \code{fitSmoothHazard}. The method
dispatch system of \proglang{R} then takes care of matching the correct
output to the correct methodology for calculating the cumulative
incidence function, without the user's intervention.

In the following sections, we illustrate these functionalities in the
context of three case studies.

@@

Our choice of modeling the log-hazard as a linear function of covariates
allows us to develop a simple computational scheme for estimation.
However, as a downside, it does not allow us to model location and scale
parameters separately like the package \pkg{flexsurv}. For example, if
we look at the Weibull distribution as parametrised in
\code{stats::pweibull}, the log-hazard function is given by
\[ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
where \(\alpha,\beta\) are shape and scale parameters, respectively.
Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
user to model the scale parameter as a function of covariates. Of
course, this added flexibility comes at the cost of interpretability: by
modeling the log-hazard directly, the parameter estimates from
\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
improve the flexibility of \pkg{casebase} at capturing the scale of a
parametric family, we could replace the logistic regression with its
quasi-likelihood counterpart and therefore model over- and
under-dispersion with respect to the logistic likelihood. We defer the
study of the properties and performance of such a model to a future
article.

@@

This report was generated on 2020-02-21 13:47:19 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 3.6.2 (2019-12-12)
#> Platform: x86_64-pc-linux-gnu (64-bit)
#> Running under: Ubuntu 18.04.4 LTS
#> 
#> Matrix products: default
#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so
#> 
#> locale:
#>  [1] LC_CTYPE=en_CA.UTF-8       LC_NUMERIC=C              
#>  [3] LC_TIME=en_CA.UTF-8        LC_COLLATE=en_CA.UTF-8    
#>  [5] LC_MONETARY=en_CA.UTF-8    LC_MESSAGES=en_CA.UTF-8   
#>  [7] LC_PAPER=en_CA.UTF-8       LC_NAME=C                 
#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
#> [11] LC_MEASUREMENT=en_CA.UTF-8 LC_IDENTIFICATION=C       
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] lubridate_1.7.4     timereg_1.9.4       glue_1.3.1         
#>  [4] cowplot_1.0.0       pracma_2.2.9        kableExtra_1.1.0   
#>  [7] data.table_1.12.8   glmnet_3.0-1        Matrix_1.2-18      
#> [10] casebase_0.2.1.9001 survival_3.1-8      magrittr_1.5       
#> [13] forcats_0.4.0       stringr_1.4.0       dplyr_0.8.3        
#> [16] purrr_0.3.3         readr_1.3.1         tidyr_1.0.0        
#> [19] tibble_2.1.3        ggplot2_3.2.1       tidyverse_1.3.0    
#> 
#> loaded via a namespace (and not attached):
#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6       
#>  [4] viridisLite_0.3.0   foreach_1.4.7       modelr_0.1.5       
#>  [7] assertthat_0.2.1    stats4_3.6.2        cellranger_1.1.0   
#> [10] yaml_2.2.0          numDeriv_2016.8-1.1 pillar_1.4.3       
#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.24      
#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.5         
#> [19] htmltools_0.4.0     pkgconfig_2.0.3     broom_0.5.2        
#> [22] haven_2.2.0         scales_1.1.0        webshot_0.5.2      
#> [25] lava_1.6.6          mgcv_1.8-31         rticles_0.12       
#> [28] generics_0.0.2      farver_2.0.3        ellipsis_0.3.0     
#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.1          
#> [34] crayon_1.3.4        readxl_1.3.1        evaluate_0.14      
#> [37] fs_1.3.1            fansi_0.4.1         MASS_7.3-51.5      
#> [40] nlme_3.1-144        xml2_1.2.2          tools_3.6.2        
#> [43] hms_0.5.2           lifecycle_0.1.0     munsell_0.5.0      
#> [46] reprex_0.3.0        compiler_3.6.2      rlang_0.4.4        
#> [49] grid_3.6.2          iterators_1.0.12    rstudioapi_0.10    
#> [52] labeling_0.3        rmarkdown_1.18      gtable_0.3.0       
#> [55] codetools_0.2-16    DBI_1.0.0           reshape2_1.4.3     
#> [58] R6_2.4.1            knitr_1.26          shape_1.4.4        
#> [61] stringi_1.4.6       Rcpp_1.0.3          vctrs_0.2.3        
#> [64] dbplyr_1.4.2        tidyselect_0.2.5    xfun_0.11
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    Jesse-Islam-master /home/turgeonm/Documents/git_repos/cbpaper
#> Head:     [3ec29a1] 2020-02-07: Merge pull request #18 from sahirbhatnagar/intro
\end{CodeOutput}
\end{CodeChunk}

>>>>

The sampling scheme selected for \code{sampleCaseBase} was described in
Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
along the ``person'' axis, proportional to each individual's total
follow-up time, and then we sample a moment uniformly over their
follow-up time. This sampling scheme is equivalent to the following
picture: imagine representing the total follow-up time of all
individuals in the study along a single dimension, where the follow-up
time of the next individual would start exactly when the follow-up time
of the previous individual ends. Then the base series could be sampled
uniformly from this one-dimensional representation of the overall
follow-up time. In any case, the output is a dataset of the same class
as the input, where each row corresponds to a person-moment. The
covariate profile for each such person-moment is retained, and an offset
term is added to the dataset. This output could then be used to fit a
smooth hazard function, or for visualization of the base series.

The fitting function \code{fitSmoothHazard} starts by looking at the
class of the dataset: if it was generated from \code{sampleCaseBase}, it
automatically inherited the class \code{cbData}. If the dataset supplied
to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
fitting function starts by calling \code{sampleCaseBase} to generate the
base series. In other words, users can bypass \code{sampleCaseBase}
altogether and only worry about the fitting function
\code{fitSmoothHazard}.

The fitting function retains the familiar formula interface of
\code{glm}. The left-hand side of the formula should be the name of the
column corresponding to the event type. The right-hand side can be any
combination of the covariates, along with an explicit functional form
for the time variable. Note that non-proportional hazard models can be
achieved at this stage by adding an interaction term involving time
(cf.~Case Study 4 below). The offset term does not need to be specified
by the user, as it is automatically added to the formula before calling
\code{glm}.

@@

\[ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
From this, we can derive the event-specific subdensities:

\[ f_j(t) = \lambda_j(t)S(t).\]

@@

\[ CI_j(t) = \int_0^t f_j(u)du.\] The integrals are computed using
either numerical integration (via the trapezoidal rule) or Monte Carlo
integration. This option is controlled by the argument \code{method} of
the \code{absoluteRisk} function.

We created \code{absoluteRisk} as an \code{S3} generic, with methods
that match the different types of outputs of \code{fitSmoothHazard}. The
method dispatch system of \proglang{R} then takes care of matching the
correct output to the correct methodology for calculating the cumulative
incidence function, without requiring any intervention from the user.

In the following sections, we illustrate these functionalities in the
context of four case studies.

@@

Our choice of modeling the log-hazard as a linear function of covariates
allows us to develop a simple computational scheme for estimation.
However, as a downside, it does not allow us to model location and scale
parameters separately like the package \pkg{flexsurv}. For example, if
we look at the Weibull distribution as parametrised in
\code{stats::pweibull}, the log-hazard function is given by
\[ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
where \(\alpha,\beta\) are shape and scale parameters, respectively.
Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
user to model the scale parameter as a function of covariates. Of
course, this added flexibility comes at the cost of interpretability: by
modeling the log-hazard directly, the parameter estimates from
\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
improve the flexibility of \pkg{casebase} at capturing the scale of a
parametric family, we could replace the logistic regression with its
quasi-likelihood counterpart and therefore model over- and
under-dispersion with respect to the logistic likelihood. We defer the
study of the properties and performance of such a model to a future
article.

@@

This report was generated on 2020-03-03 13:34:41 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 3.6.2 (2019-12-12)
#> Platform: x86_64-pc-linux-gnu (64-bit)
#> Running under: Ubuntu 18.04.4 LTS
#> 
#> Matrix products: default
#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so
#> 
#> locale:
#>  [1] LC_CTYPE=en_CA.UTF-8       LC_NUMERIC=C              
#>  [3] LC_TIME=en_CA.UTF-8        LC_COLLATE=en_CA.UTF-8    
#>  [5] LC_MONETARY=en_CA.UTF-8    LC_MESSAGES=en_CA.UTF-8   
#>  [7] LC_PAPER=en_CA.UTF-8       LC_NAME=C                 
#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
#> [11] LC_MEASUREMENT=en_CA.UTF-8 LC_IDENTIFICATION=C       
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] lubridate_1.7.4     timereg_1.9.4       glue_1.3.1         
#>  [4] cowplot_1.0.0       pracma_2.2.9        kableExtra_1.1.0   
#>  [7] data.table_1.12.8   glmnet_3.0-1        Matrix_1.2-18      
#> [10] casebase_0.2.1.9001 survival_3.1-8      magrittr_1.5       
#> [13] forcats_0.4.0       stringr_1.4.0       dplyr_0.8.3        
#> [16] purrr_0.3.3         readr_1.3.1         tidyr_1.0.0        
#> [19] tibble_2.1.3        ggplot2_3.2.1       tidyverse_1.3.0    
#> 
#> loaded via a namespace (and not attached):
#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6.1     
#>  [4] viridisLite_0.3.0   foreach_1.4.7       modelr_0.1.5       
#>  [7] assertthat_0.2.1    stats4_3.6.2        cellranger_1.1.0   
#> [10] yaml_2.2.1          numDeriv_2016.8-1.1 pillar_1.4.3       
#> [13] backports_1.1.5     lattice_0.20-40     digest_0.6.25      
#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.6         
#> [19] htmltools_0.4.0     pkgconfig_2.0.3     broom_0.5.2        
#> [22] haven_2.2.0         scales_1.1.0        webshot_0.5.2      
#> [25] lava_1.6.6          mgcv_1.8-31         rticles_0.12       
#> [28] generics_0.0.2      farver_2.0.3        ellipsis_0.3.0     
#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.2          
#> [34] crayon_1.3.4        readxl_1.3.1        evaluate_0.14      
#> [37] fs_1.3.1            fansi_0.4.1         MASS_7.3-51.5      
#> [40] nlme_3.1-144        xml2_1.2.2          tools_3.6.2        
#> [43] hms_0.5.2           lifecycle_0.1.0     munsell_0.5.0      
#> [46] reprex_0.3.0        compiler_3.6.2      rlang_0.4.5        
#> [49] grid_3.6.2          iterators_1.0.12    rstudioapi_0.10    
#> [52] labeling_0.3        rmarkdown_1.18      gtable_0.3.0       
#> [55] codetools_0.2-16    DBI_1.0.0           reshape2_1.4.3     
#> [58] R6_2.4.1            knitr_1.26          shape_1.4.4        
#> [61] stringi_1.4.6       Rcpp_1.0.3          vctrs_0.2.3        
#> [64] dbplyr_1.4.2        tidyselect_0.2.5    xfun_0.11
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    review /home/turgeonm/Documents/git_repos/cbpaper
#> Head:     [96bfdb4] 2020-03-03: expand theoretical details
\end{CodeOutput}
\end{CodeChunk}
